---
title: "Income Level and Access to Public Transportation"
author: Eric Forteza
description: How does income level affect access to public transportation in the United States?
output:
  distill::distill_article:
    self_contained: false
---




## Introduction

Does a person's income level affect their access to public transportation? In this study, I plan to examine the extent to which income level influences how much access people have to public transportation. I hypothesize that people with higher incomes will people with higher incomes will have more access to public transportation infrastructure than those with lower incomes. Often, higher incomes allow local government to invest in public infrastructure, like transportation, through taxation. However, it is also often true that people with higher incomes tend to live in more suburban areas, which have less public transportation. This project seeks to offer more clarity to this puzzle by exploring how wealth in the US relates to access to public transportation. Specifically: our project asks whether the income level of a block group in the US influences its proximity to a public transportation stop. If I observe that block groups with higher percentages of low wage workers have less proximity to public transportation stops, this would support my hypothesis. However, if I observe that block groups with lower percentages of low wage workers have more proximity or a similar proximity to public transportation stops, this would provide eveidence against my hypothesis. In addition, this project will look at potential confounders variable like density, frequency of public transportation during peak hours, and racial diversity to have a more clear understanding of this relationship. 

## Data and Research Design


To investigate this relationship, I am relying on data from the following source:

United States Environmental Protection Agency's National Walkability Index (2021). This nationwide geographic data source ranks block groups according to their relative walkability. This data set includes the block groups' walkability scores as well as underlying attributes that are used to rank them, including the income levels of block groups, the percentage of low-income individuals per block, proximity to transit, etc. 

Our analyses cover percentage of low-income individuals and proximity to transit stop in meters for 84,912 unique block groups from the 2019 US Census. My research design is observational, correlational analysis, and makes no casual assumptions or claims. 

```{r}

library (tidyverse) 
library (modelsummary)
library(kableExtra)
library(broom)
```
```{r}
options(
  knitr.kable.NA = '',
  width = 100)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center")
```

```{r}
library (tidyverse)
walkability_score <- read.csv ("EPA_SmartLocationDatabase_V3_Jan_2021_Final.csv")

```



```{r}
walkability_score <- walkability_score |> 
  mutate (`Proximity to Transportation Stop`= D4A,
          `Percentage of Low Wage Workers` = R_PCTLOWWAGE,
          `Population Density` = D1B,
          `Auto-oriented Road Density` = D3AAO,
            `Multi-Modal Road Density` = D3AMM,
          `Pedestrian-Oriented Road Density` = D3APO)
  
income_and_transportation_access <- walkability_score |>
  filter(`Proximity to Transportation Stop` >= 0) |>
  na.omit (`Proximity to Transportation Stop`) |>
  mutate (`Proximity to Transportation Stop` = `Proximity to Transportation Stop` / 1609) 


```

National Walkability Index provides us with our outcome variable: the proximity of a block-group to a public transportation stop in miles. This was originally measured in meters from population-weighted centroid of block group to nearest transit stop; however, I have changed the scale from meters to miles. The outcome ranges from 0 miles to .75 miles, with a median of 0.2451 and a mean of 0.2791. 


```{r}
transit_prox <- income_and_transportation_access |>
  select (`Proximity to Transportation Stop`) |>
  summary()

transit_prox

transit_prox_lineplot<- income_and_transportation_access |>
  ggplot (mapping = aes (x= `Proximity to Transportation Stop`)) +
  geom_histogram ( color = "white") +
  labs(x = "Proximity to a Transportation Stop from Home in Miles",
        y = "Count",
        title = "Distribution of Proximity to a Transportation Stop from Home") 

transit_prox_lineplot
```


National Walkability Index also provides us with our explanatory variable: the percent of low wage workers in a block-group based on home location, with low wage workes being defined as workers who make $1250 per month or less. This data set classifies medium wage workers as those making more than $1250 but less than $3333 and high wage works as those making more than $3333. This variable has a median of 0.226, and a mean of 0.232.

```{r}
income_level <- income_and_transportation_access |>
  select (`Percentage of Low Wage Workers`) |>
  summary()

income_level



income_level_lineplot<- income_and_transportation_access |>
  ggplot (mapping = aes (x= `Percentage of Low Wage Workers`)) +
  geom_histogram (bandwidth = 10,  color = "white") +
  labs(x = "Percent of Low Wage Workers Per Block Group",
        y = "Count",
        title = "Distribution of Percent of Low Wage Workers Per Block Group") 

income_level_lineplot
```

## Results


Before regressing the proximity to a transportation stop on the percentage of low wage workers in a block group, the bivarate relationship between these variables can be visualized with a scatter plot. The scatter plot graph below plots the point percentage of low wage workers per block group on the x-axis, and the prxomitiy in miles to a public transportation stop.




```{r}
ggplot(data = income_and_transportation_access, mapping = aes(x = `Percentage of Low Wage Workers`, y = `Proximity to Transportation Stop`)) + 
  geom_point(alpha = 0.2) +
  geom_smooth (method = 'lm',  se = FALSE, size = 1) +
  labs (title = 
  "Proximity of Public Transportation Stop by 
   Percentage of Low Wage Workers per Block Group") +
  theme(plot.title = element_text(hjust = 0.5)) +
  xlim(0, 1) 

```
The bivariate scatter plot suggests that there exists a negative relationship between higher percentage of low wage workers in a block group, and the distance to a public transportation stop. A regression analysis will help to provide a more detailed picture about this seemingly negative association.
  
**Regression Results**

Below you can find the results from two regression models. The first is a bivariate model, which regresses the proximity to a transportation stop on the percentage of low wage workers in a block group. The second is a multivariate model which includes block groun level covariates such as population density, the network density in terms of facility miles of auto-oriented links per square miles, network density in terms of facility miles of multi-modal links per square mile, and network density in terms of facility miles of pedestrian-oriented links per square mile

The regression results form both models show that a statistically significant, negative correlation exists between the percentage of low wage workers per block group and proximity to a public transportation stop. 

```{r}
model_1 <- lm(
  data = income_and_transportation_access, 
  `Proximity to Transportation Stop` ~ `Percentage of Low Wage Workers`)

model_2 <- lm(
  data = income_and_transportation_access, 
  `Proximity to Transportation Stop` ~ `Percentage of Low Wage Workers` + `Population Density` + `Auto-oriented Road Density` + `Multi-Modal Road Density` + `Pedestrian-Oriented Road Density`) 


results <-
  full_join(
    model_1 |> 
      broom::tidy() |> 
      select(term, est1 = estimate, se1 = std.error, pval1 = p.value),
    model_2 |> 
      broom::tidy() |> 
      select(term, est2 = estimate, se2 = std.error, pval2 = p.value)) |>
  select(term, matches("est"), matches("se"), matches ("pval")) 


results |>
  knitr::kable(
    booktabs = TRUE, digits = 4,
    col.names = c("Coefficient", rep(c("Bivariate", "Multivariate"), 3))) |>
  kable_styling(
    position = "center", 
    latex_options = "HOLD_position") |>
   add_header_above( c(
      " " = 1, 
      "Coef. estimate" = 2, 
      "Std. error" = 2, 
      "P-value" = 2)) |>
  column_spec(1, color = "black", background = "white")
```

The bivariate model suggests that on average, when the percentage of low wage workers in a block group is 0, the distance to a public transportation stop is .349 miles. The coefficient on the percentage of low-wage workers is -0.299, which means that when there is a one unit increase in the percentage of low-wage workers from a scale of 0 to 1, distance to a public transportation stop is predicted to be about .299 miles closer on average. 

The multivariate model predicts a larger association. Holding population density, the network density in terms of facility miles of auto-oriented links per square miles, network density in terms of facility miles of multi-modal links per square mile, and network density in terms of facility miles of pedestrian-oriented links per square mile constant, a one unit increase in the proportion of low wage workers is associated with a decrease in proximity to a public transportation stop by .362 miles.  

**Statistical Significance**


The association between the proximity to public transportation and the percentage of low wage workers per block group is statistically significant at the test level α = 0.05 in both the bivariate amd multivariate models. In both models, the p-value on the coefficient is 0. Since the p-value is 0, we would reject the null hypothesis that there is no association between the proximity to public transportation and the percentage of low wage workers per block group. In other words, the hypothesis that the association between the proximity to public transportation and the percentage of low wage workers per block group was non-zero, then the coefficient's p-value of 0 would suggest that there is a near 100% probability that the estimated association, in this case a negative association, is not due to random chance. 



```{r}
#Coefficient estimates
coefest_model_1 <- summary(model_1)$coefficients["`Percentage of Low Wage Workers`", "Estimate"]
coefest_model_2 <- summary(model_2)$coefficients ["`Percentage of Low Wage Workers`", "Estimate"]
#Standard errors
coefse_model_1 <- summary(model_1)$coefficients ["`Percentage of Low Wage Workers`", "Std. Error"]
coefse_model_2 <- summary(model_2)$coefficients["`Percentage of Low Wage Workers`", "Std. Error"]
#Probabilistic onfidence intervals
ci95_model_1 <- c(
  coefest_model_1 -1.96*coefse_model_1,
  coefest_model_1 + 1.96*coefse_model_1)
ci95_model_2 <- c(
  coefest_model_2 - 1.96*coefse_model_2,
  coefest_model_2 + 1.96*coefse_model_2)

ci95_model_1

ci95_model_2
```

Assuming the sampling distribution of the proximity to public transportation is approximately normal, the standard error on the coefficient estimates for the percentage of low wage workers per block group cna be used to construct a probabilistic 95% confidence interval. Since 1.96 represents the z-score that captures 95% of the data underneath a standard normal distirbution, that value can be multiplied by the coefficient standard error, and this product subtracted and added to the coefficient point estimate, to produce lower and upper bound limits of the 95% confidence interval.

The 95% confidence interval for ideal point preferences under the bivariate model can be formally described as

$$
\beta_{\text{`Percentage of Low Wage Workers`}} =-0.2993	 \in  (-0.318,\,-0.279)
$$
and under the multivariate model as

$$
\beta_{\text{`Percentage of Low Wage Workers`}} =-0.3616		 \in  (-0.379,\,-0.343)
$$

This means that across repeated samples, 95% of regression estimated correlations between the proximity to public transportation and the percentage of low wage workers per block group would be within the bounds above. 
