---
title: "Income Level and Access to Public Transportation"
author: Eric Forteza
description: How does income level affect access to public transportation in the United States?
output:
  distill::distill_article:
    self_contained: false
---




## Introduction

Does a person's income level affect their access to public transportation? In this study, I plan to examine the extent to which income level influences how much access people have to public transportation. I hypothesize that people with higher incomes will have more access to public transportation infrastructure than those with lower incomes. Often, higher incomes allow local government to invest in public infrastructure, like transportation, through taxation. Additionally, many urban areas with more developed public transportation infrastructure are often very expensive to live in, such as New York City, Chicago, Boston, etc. However, it is also often true that people with higher incomes tend to live in more suburban areas, which have less public transportation. This project seeks to offer more clarity to this puzzle by exploring how wealth in the US relates to access to public transportation. Specifically: our project asks whether the income level of a block group in the US influences its proximity to a public transportation stop. If I observe that block groups with higher percentages of low wage workers are further away from public transportation stops, this would support my hypothesis. However, if I observe that block groups with higher percentages of low wage workers are actually closer to public transportation stops, this would provide evidence against my hypothesis. 

## Data and Research Design

This project is relying on the following data source to investigate this relationship:

United States Environmental Protection Agency's National Walkability Index (2021). This nationwide geographic data source ranks block groups according to their relative walkability. This data set includes the block groups' walkability scores as well as underlying attributes that are used to rank them, including the income levels of block groups, the percentage of low-income individuals per block, proximity to transit, etc. 

This analyses covers percentage of low-income individuals and proximity to transit stop in meters for 84,912 unique block groups from the 2019 US Census. My research design is observational, correlational analysis, and makes no casual assumptions or claims. 

```{r}
library (tidyverse) 
library (modelsummary)
library(kableExtra)
library(broom)
```
```{r}
options(
  knitr.kable.NA = '',
  width = 100)
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center")
```

```{r}
library (tidyverse)
walkability_score <- read.csv ("EPA_SmartLocationDatabase_V3_Jan_2021_Final.csv")

```



```{r}
walkability_score <- walkability_score |> 
  mutate (`Proximity to Transportation Stop`= D4A,
          `Percentage of Low Wage Workers` = R_PCTLOWWAGE,
          `Population Density` = D1B,
          `Auto-oriented Road Density` = D3AAO,
            `Multi-Modal Road Density` = D3AMM,
          `Pedestrian-Oriented Road Density` = D3APO)
  
income_and_transportation_access <- walkability_score |>
  filter(`Proximity to Transportation Stop` >= 0) |>
  na.omit (`Proximity to Transportation Stop`) |>
  mutate (`Proximity to Transportation Stop` = `Proximity to Transportation Stop` / 1609) 


```

National Walkability Index provides us with our outcome variable: the proximity of a block-group to a public transportation stop in miles. This was originally measured in meters from population-weighted centroid of block group to nearest transit stop; however, I have changed the scale from meters to miles to create a 0 to 1 scale that would be simpler to analyze. The outcome ranges from 0 miles to .75 miles, with a median of 0.2451 and a mean of 0.2791 miles. 


```{r}
transit_prox <- income_and_transportation_access |>
  select (`Proximity to Transportation Stop`) |>
  summary()

transit_prox

transit_prox_lineplot<- income_and_transportation_access |>
  ggplot (mapping = aes (x= `Proximity to Transportation Stop`)) +
  geom_histogram ( color = "white") +
  labs(x = "Proximity to a Transportation Stop from Home in Miles",
        y = "Count",
        title = "Distribution of Proximity to a Transportation Stop from Home") 

transit_prox_lineplot
```


National Walkability Index also provides us with our explanatory variable: the percent of low wage workers in a block-group based on home location, with low wage workers being defined as workers who make $1250 per month or less. This data set classifies medium wage workers as those making more than $1250 but less than $3333 and high wage works as those making more than $3333. This variable has a median of 0.226, and a mean of 0.232.

```{r}
income_level <- income_and_transportation_access |>
  select (`Percentage of Low Wage Workers`) |>
  summary()

income_level



income_level_lineplot<- income_and_transportation_access |>
  ggplot (mapping = aes (x= `Percentage of Low Wage Workers`)) +
  geom_histogram (bandwidth = 10,  color = "white") +
  labs(x = "Percent of Low Wage Workers Per Block Group",
        y = "Count",
        title = "Distribution of Percent of Low Wage Workers Per Block Group") 

income_level_lineplot
```

## Results

Prior to performing a regression of the proximity to a transportation stop on the percentage of low wage workers in a block group, a scatter plot can visualize the bivariate relationship between these two variables. The scatter plot graph below plots the point percentage of low wage workers per block group on the x-axis, and the proximity in miles to a public transportation stop.

```{r}
ggplot(data = income_and_transportation_access, mapping = aes(x = `Percentage of Low Wage Workers`, y = `Proximity to Transportation Stop`)) + 
  geom_point(alpha = 0.2) +
  geom_smooth (method = 'lm',  se = FALSE, size = 1) +
  labs (title = 
  "Proximity of Public Transportation Stop by 
   Percentage of Low Wage Workers per Block Group") +
  theme(plot.title = element_text(hjust = 0.5)) 

```
The bivariate scatter plot suggests that there exists a negative relationship between higher percentage of low wage workers in a block group, and the distance to a public transportation stop. A regression analysis will help to provide a more detailed picture about this seemingly negative association.
  
**Regression Results**

Below   can find the results from two regression models. The first is a bivariate model, which regresses the proximity to a transportation stop on the percentage of low wage workers in a block group. The second is a multivariate model which includes block group level covariates such as population density, the network density in terms of facility miles of auto-oriented links per square miles, network density in terms of facility miles of multi-modal links per square mile, and network density in terms of facility miles of pedestrian-oriented links per square mile

The regression results form both models show that a statistically significant, negative correlation exists between the percentage of low wage workers per block group and proximity to a public transportation stop. 

```{r}
model_1 <- lm(
  data = income_and_transportation_access, 
  `Proximity to Transportation Stop` ~ `Percentage of Low Wage Workers`)

model_2 <- lm(
  data = income_and_transportation_access, 
  `Proximity to Transportation Stop` ~ `Percentage of Low Wage Workers` + `Population Density` + `Auto-oriented Road Density` + `Multi-Modal Road Density` + `Pedestrian-Oriented Road Density`) 


results <-
  full_join(
    model_1 |> 
      broom::tidy() |> 
      select(term, est1 = estimate, se1 = std.error, pval1 = p.value),
    model_2 |> 
      broom::tidy() |> 
      select(term, est2 = estimate, se2 = std.error, pval2 = p.value)) |>
  select(term, matches("est"), matches("se"), matches ("pval")) 


results |>
  knitr::kable(
    booktabs = TRUE, digits = 4,
    col.names = c("Coefficient", rep(c("Bivariate", "Multivariate"), 3))) |>
  kable_styling(
    position = "center", 
    latex_options = "HOLD_position") |>
   add_header_above( c(
      " " = 1, 
      "Coef. estimate" = 2, 
      "Std. error" = 2, 
      "P-value" = 2)) |>
  column_spec(1, color = "black", background = "white")
```

The bivariate model suggests that on average, when the percentage of low wage workers in a block group is 0, the distance to a public transportation stop is .349 miles. The coefficient on the percentage of low-wage workers is -0.299, which means that when there is a one unit increase in the percentage of low-wage workers from a scale of 0 to 1, distance to a public transportation stop is predicted to be about .299 miles closer on average. 

The multivariate model predicts a larger association. Holding population density, the network density in terms of facility miles of auto-oriented links per square miles, network density in terms of facility miles of multi-modal links per square mile, and network density in terms of facility miles of pedestrian-oriented links per square mile constant, a one unit increase in the proportion of low wage workers is associated with a decrease in proximity to a public transportation stop by .362 miles.  

**Statistical Significance**


The association between the proximity to public transportation and the percentage of low wage workers per block group is statistically significant at the test level α = 0.05 in both the bivariate and multivariate models. In both models, the p-value on the coefficient is 0. Since the p-value is 0, we would reject the null hypothesis that there is no association between the proximity to public transportation and the percentage of low wage workers per block group. In other words, the hypothesis that the association between the proximity to public transportation and the percentage of low wage workers per block group was non-zero, then the coefficient's p-value of 0 would suggest that there is a near 100% probability that the estimated association, in this case a negative association, is not due to random chance. 



```{r}
#Coefficient estimates
coefest_model_1 <- summary(model_1)$coefficients["`Percentage of Low Wage Workers`", "Estimate"]
coefest_model_2 <- summary(model_2)$coefficients ["`Percentage of Low Wage Workers`", "Estimate"]
#Standard errors
coefse_model_1 <- summary(model_1)$coefficients ["`Percentage of Low Wage Workers`", "Std. Error"]
coefse_model_2 <- summary(model_2)$coefficients["`Percentage of Low Wage Workers`", "Std. Error"]
#Probabilistic onfidence intervals
ci95_model_1 <- c(
  coefest_model_1 -1.96*coefse_model_1,
  coefest_model_1 + 1.96*coefse_model_1)
ci95_model_2 <- c(
  coefest_model_2 - 1.96*coefse_model_2,
  coefest_model_2 + 1.96*coefse_model_2)

ci95_model_1

ci95_model_2
```

Assuming the sampling distribution of the proximity to public transportation is approximately normal, the standard error on the coefficient estimates for the percentage of low wage workers per block group can be used to construct a probabilistic 95% confidence interval. Since 1.96 represents the z-score that captures 95% of the data underneath a standard normal distribution, that value can be multiplied by the coefficient standard error, and this product subtracted and added to the coefficient point estimate, to produce lower and upper bound limits of the 95% confidence interval.

The 95% confidence interval for ideal point preferences under the bivariate model can be formally described as

$$
\beta_{\text{`Percentage of Low Wage Workers`}} =-0.2993	 \in  (-0.318,\,-0.279)
$$
and under the multivariate model as

$$
\beta_{\text{`Percentage of Low Wage Workers`}} =-0.3616		 \in  (-0.379,\,-0.343)
$$

This means that across repeated samples, 95% of regression estimated correlations between the proximity to public transportation and the percentage of low wage workers per block group would be within the bounds above. 


**Causality and confounders**

Since the explanatory variable of percentage of low wage workers per block group is not randomly assigned, but is rather a result of years of changes and developments in each block group, it cannot be treated as a "treatment" variable. Therefore, increases in the percentage of  low wage workers per block group cannot be said to casually decrease the proximity to a public transportation stop. Instead, we can say that the two phenomenon are negatively associated with each other, with this estimated association being statistically significant. 

The multivariate model attempts to control for various confounders. One of these is the population density of the block groups. Since higher population densities, especially in urban areas, are often associated with more investment in public transportation and therefore more access to buses, subways, etc, this could affect the proximity of each block to the nearest transit stop. This multivariate model also attempts to account for the network density of the block groups in terms of the facility of traveling using auto-oriented infrastructure, pedestrian-oriented infrastructure, or a mixture of both. Since these measurements help to showcase what kind of traveling infrastructure is available on each block group, it can affect the distance to the nearest transit stop. For example, block grouns with a higher network density for cars might have longer distances to the nearest transit stops because the city or town it is located likely invested more on creating car-friendly infrastructure. Accounting for these potential confounders help to create a more objective understanding of the relationship between  the proximity to public transportation and the percentage of low wage workers per block group.

Other possible confounders exist, but are excluded from this  model due to data unavailability. For example, it would be useful to have further categories of public transportation stops. Often, bus stops are significantly cheaper to construct and maintain than subway or metro stops. Having the classification for various types of public transportation stops would provide further understanding of the the relationship between income level and access to public transportation. Another potential confounders is the amount of tax revenue that each block group generates. Since most of public transportation funding comes from local and state taxes, accounting for how much taxes different block groups pay, which would likely differ based on their income levels, would provide a helpful understanding of how wealth affects access top public transportation by understanding how much funding each city has to spend on transportation infrastructure. This could prove even more helpful if we could find data on the total spending of each block group on public transportation infrastructure, since it would have a significant effect on this transportation network and therefore proximity.   

**Conclusion and Future Steps**

The regression results in this study contradict my original hypothesis that people with higher income levels will have more access to public transportation than those with lower income levels. Instead, this study's bivariate model demonstrate that block groups with lower levels of low wage workers are negatively and significantly associated with closer proximity to public transportation stops. The multivariate model that controls for possible influential confounders and covariates, including  population density, the network density in terms of facility miles of auto-oriented links per square miles, network density in terms of facility miles of multi-modal links per square mile, and network density in terms of facility miles of pedestrian-oriented links per square mile, demonstrate an even larger estimate of this association.  

This analysis is limited by having a partially incomplete sample from the population. Although the EPA's National Walkability Index accounts for most if not all block grounds in the US, almost 2/3 of these did not have data for the proximity to public transportation stops. Another limitation is how this study determines income levels. Currently,  low wage workers are classified as those who make $1250 per month or less, medium wage workers are those making mroe than $1250 but less than $3333, and high wage workers are classified as those making more than $3333. The main issue with this classification is that the salary range for high wage workers is too spread out, as someone making $39,996 a year would be in the same category as someone making $100,000 or even 1,000,000.  

One way to improve this study in the future is by creating new classifications of income levels as a way to understand the effect that this variable has on public transportation access.
Using the Pew Research Center's data on median incomes for lower, middle, and upper incomes, one could create the following classifications: lower income has those making less than $29,963 per year, middle income has those making more than $29,963 and less than $90,131 a year, and upper income has those making more than $90,131 a year. This would likely yyield a more accurate measurement of the relationship between welath and access to public transportation. Additionally, future studies should also look at how income level affects the frequency and logistics of using public transportation as a means to travel. This would provide a truer measurement of access, rather than just proximity. 